How does character divergence affect resource dynamics?

### 2018-05-11 ECD_Supp_Mat.Rmd
adding stability results

---
title: "How does character divergence affect resource dynamics?"
output: html_document
author: Matt Barbour
---

@@

$$\frac{dR_1}{dt}=r_1K_1(1-\frac{R_1}{K_1})-F_{11}(R_1)C_1-F_{21}(R_1)C_2$$
$$\frac{dR_2}{dt}=r_2K_2(1-\frac{R_2}{K_2})-F_{12}(R_2)C_1-F_{22}(R_2)C_2$$
$$\frac{dC_1}{dt}=e_{11}F_{11}(R_1)C_1+e_{12}F_{12}(R_2)C_1-m_1C_1$$
$$\frac{dC_2}{dt}=e_{21}F_{21}(R_1)C_2+e_{22}F_{22}(R_2)C_2-m_2C_2$$
where $r_i$ represents intrinsic growth rates, $K_i$ represents carrying capacity, $e_{ij}$ represents the conversion efficiency of resources into new consumers, and $m_i$ represents the mortality rate of consumers. $F_{ij}(R_j)$ represents the consumer's feeding rate (or functional response) on a specific resource (in this case, $C_i$ on $R_j$). This model is a useful characterization of a scenario where consumers compete for two distinct resources (e.g. zooplankton and benthic invertebrates) rather than if resources are better characterized by a continuous trait distribution (e.g. seed size; Taper and Case 1985). In the scenario where consumer feeding rates increase linearly with resource density as a function of their attack rate, such that $F_{ij}(R_j)=a_{ij}R_j$, then our model becomes MacArthur's model of resource competition (MacArthur 1972). The evolution of consumer attack rates in this model, and several extensions, have been analyzed in detail (Lawlor and Smith 1976; Abrams 1986). The general result of these analyses is that consumer divergence is an inevitable outcome of competition. Here, we examine the effects of this inevitable divergence on resource dynamics and food-web stability. We show that the ecological consequences of character displacement are not intuitive and depend critically on the foraging scenario and tradeoff in consumer attack rates.

@@

The first foragine scenario we explore is already assumed in MacArthur's model of resource competition, namely that consumers can forage for both resources simultaneously. This also implies that both resources occur within the same microhabitat. To gain analytical insight to this scenario, we assume that resources are equivalent ($r=r_i$ and $K=K_i$) as well as consumers ($e=e_{ii}=e_{ij}$; $m=m_i$), except that consumer attack rates are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$). While this scenario is simple, it isolates the effects of character displacement, rather than any inherent differences in consumers or resources. In the appendix, we show that when both consumers and resources are present that the abundance of resources at equilibrium are equivalent ($R_i=R_1=R_2$) and are determined by the following equation:

$$\hat{R_i}=\frac{1}{a_{ii}+a_{ij}}\cdot\frac{m}{e}$$
Note that a key determinant of resource abundance is the consumer's total attack rate, $a_{ii}+a_{ij}$. Since the total attack rate is in the denominator, this implies that resource abundances will decrease if evolution favors an increase in total attack rate. Consumers, however, cannot maximize their attack rate

To understand how a consumer's total attack rate would change with character divergence, we explored the effect of different tradeoffs in consumer attack rates. We used the following general function:

$$\Big(\frac{a_{ii}}{A}\Big)^n+\Big(\frac{a_{ij}}{A}\Big)^n=1$$

where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff. This function has the useful property that it differentiates between cases where intermediate combinations of $a_{ii}$ and $a_{ij}$ are higher, on average, than the extremes (when $n>1$) or, conversely, where the two extremes are higher, on average, than intermediate investments (when $n<1$; Sargent and Otto 2006). When $n=1$, the tradeoff function is linear, and all combinations of $a_ii$ and $a_ij$ have the same total attack rate.

```{r Tradeoff function, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
bind_rows(convex_df, linear_df, concave_df) %>%
  ggplot(., aes(x=aii, y=aij, linetype = factor(n))) + 
  geom_line() + 
  xlab(expression(a[ii])) +
  ylab(expression(a[ij])) +
  scale_linetype_manual(values=1:3, name="Tradeoff (n)") 
```

DESCRIBE THE SIMULATIONS THAT ARE USED IN ALL OF THE FOLLOWING ANALYSES.

@@

```{r Same resource habitat, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
Mac_attack <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=aii+aij, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=a11+a12, fill=Competitor), shape=21, size=5, inherit.aes = FALSE) +
  ylab(expression(a[ii]+a[ij])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_linetype_manual(values=1:3, name="Tradeoff (n)") +
  scale_fill_manual(values=c("white","black"))

@@

Mac_plots <- plot_grid(Mac_attack + theme(legend.position = "none"), Mac_resources + theme(legend.position = "none"), nrow=1)

@@

Mac_plot_legend <- plot_grid(Mac_plots, Mac_legend, ncol=2, rel_widths = c(3,0.3), align='hv')

Mac_title <- ggdraw() + draw_label("Resources in same habitat", fontface='bold')
plot_grid(Mac_title, Mac_plot_legend, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins
```

@@

```{r Distinct resource habitats, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
LS_attack <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=wii*aii+wij*aij, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=w11*a11+w12*a12, fill=Competitor), shape=21, size=5, inherit.aes = FALSE) +
  ylab(expression(w[ii]*a[ii]+(1-w[ii])*a[ij])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_linetype_manual(values=1:3, name="Tradeoff (n)") +
  scale_fill_manual(values=c("white","black"))

@@

LS_plots <- plot_grid(LS_attack + theme(legend.position = "none"), LS_resources + theme(legend.position = "none"), nrow=1)

@@

LS_plot_legend <- plot_grid(LS_plots, LS_legend, nrow=1, rel_widths = c(3,0.3))

LS_title <- ggdraw() + draw_label("Resources in different habitats", fontface='bold')
plot_grid(LS_title, LS_plot_legend, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins
```

@@

```{r}
McCann_attack <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=wii*aii+wij*aij, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(McCann_displacements, Time=="End"), aes(x = a11/(a11+a12), y=w11*a11+w12*a12, fill=Competitor), shape=21, size=5, inherit.aes = FALSE) +
  ylab(expression(w[ii]*a[ii]+(1-w[ii])*a[ij])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_linetype_manual(values=1:3, name="Tradeoff (n)") +
  scale_fill_manual(values=c("white","black"))

@@

McCann_plots <- plot_grid(McCann_attack + theme(legend.position = "none"), McCann_resources + theme(legend.position = "none"), nrow=1)

@@

McCann_plot_legend <- plot_grid(McCann_plots, McCann_legend, nrow=1, rel_widths = c(3,0.3))

McCann_title <- ggdraw() + draw_label("Resources in different habitats & realistic functional response", fontface='bold')
plot_grid(McCann_title, McCann_plot_legend, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins
```

@@

This relationship between character divergence and resource abundances begs the question: how does character divergence affect food-web stability? This question has been addressed empirically by Pimentel as well as with a conceptual model. Lawlor and Smith though did a detailed theoretical analysis of how character divergence affects three different aspects of stability. They looked at stability through the lens of competition. The first aspect was whether character divergence resulted in mutual invasibility of consumers, a common metric of "stable coexistence" in competition models. Another very interesting aspect they looked at was a global metric that examined the persistence of the system to perturbations in resources. Here, they conclude the divergence increases community stability, because it expands the range of resource perturbations that a system can withstand before a consumer going extinct. Note though, that they overlooked that there conclusions do not actually quantify the effects of ecological character displacement on stability because they do not compare scenarios where a consumer is in allopatry and one where consumers are in sympatry. Finally, they also look at the Lyapunov exponent (local stability), and while they find a negative effect of divergence, they suggest that it would likely be overshadowed by the stabilizing effects of divergence on mutual invasibility and global stability.

>>>>

---
title: "How does character divergence affect resource dynamics?"
output: html_document
  fig_caption: yes
author: Matt Barbour
---

@@

$$\frac{dR_1}{dt}=r_1K_1(1-\frac{R_1}{K_1})-F_{11}(R_1)C_1-F_{21}(R_1)C_2$$
$$\frac{dR_2}{dt}=r_2K_2(1-\frac{R_2}{K_2})-F_{12}(R_2)C_1-F_{22}(R_2)C_2$$
$$\frac{dC_1}{dt}=e_{11}F_{11}(R_1)C_1+e_{12}F_{12}(R_2)C_1-m_1C_1$$
$$\frac{dC_2}{dt}=e_{21}F_{21}(R_1)C_2+e_{22}F_{22}(R_2)C_2-m_2C_2$$
Resource growth is determined by is intrinsic growth rate ($r_i$), carrying capacity ($K_i$), and the feeding rate (or functional response) of each consumer ($F_{ii}(R_i)$, $F_{ij}(R_j)$). Consumer growth is determined by its feeding rate on each resource as well as its conversion efficiency ($e_{ij}$) and mortality rate ($m_i$). This model is a useful characterization of a scenario where consumers compete for two distinct resources (e.g. zooplankton and benthic invertebrates) rather than if resources are better characterized by a continuous trait distribution (e.g. seed size; Taper and Case 1985). In the scenario where consumer feeding rates increase linearly with resource density as a function of their attack rate ($a_{ij}$), such that $F_{ij}(R_j)=a_{ij}R_j$, then our model becomes MacArthur's model of resource competition (MacArthur 1972). The evolution of consumer attack rates in this model, and several extensions, have been analyzed in detail (Lawlor and Smith 1976; Abrams 1986). The general result of these analyses is that consumer divergence is an inevitable outcome of resource competition.

Here, we examine the effects of this inevitable divergence on resource dynamics.

GIVE DETAILS ABOUT MY ANALYTICAL ANALYSIS AS WELL AS THE TYPE OF SIMULATIONS I PERFORM.

We show that the ecological consequences of character displacement are not intuitive and depend critically on the foraging scenario and tradeoff in consumer attack rates.

@@

The first foraging scenario we examine is already assumed in MacArthur's model of resource competition, namely that consumers can forage for both resources simultaneously (e.g. they occur in the same habitat). To gain analytical insight to this scenario, we assume that resources are equivalent ($r=r_i$ and $K=K_i$) as well as consumers ($e=e_{ii}=e_{ij}$; $m=m_i$), except that consumer attack rates are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$). While this scenario is simple, it allow us to isolate the effects of character displacement rather than any inherent differences in consumers or resources. In the appendix, we show that when both consumers and resources are present, the abundance of resources at equilibrium are equivalent ($R=R_1=R_2$) and are determined by the following equation:

$$\hat{R}=\frac{1}{a_{ii}+a_{ij}}\cdot\frac{m}{e}$$
Note that a key determinant of resource abundance is the consumer's total attack rate, $a_{ii}+a_{ij}$. Since the total attack rate is in the denominator, this implies that resource abundances will decrease if consumers evolve higher total attack rates. If there were no constraints on attack rates, each consumer would evolve increasingly higher attack rates on each resource until it goes extinct. In consideration of this, we have included a tradeoff between investments in attacking one resource versus the other such that $\big(a_{ii}/{A}\big)^n+\big(a_{ij}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff. This function has the useful property that it differentiates between cases where intermediate combinations of $a_{ii}$ and $a_{ij}$ are higher, on average, than the extremes (when $n>1$) or, conversely, where the two extremes are higher, on average, than intermediate investments (when $n<1$; Sargent and Otto 2006). When $n=1$, the tradeoff function is linear, and all combinations of $a_ii$ and $a_ij$ have the same total attack rate.

```{r Tradeoff function, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.cap="Tradeoffs in consumer attack rates"}
bind_rows(concaveDN_df, linear_df, concaveUP_df) %>%
  ggplot(., aes(x=aii, y=aij, linetype = factor(n))) + 
  geom_line() + 
  xlab(expression(a[ii])) +
  ylab(expression(a[ij])) +
  scale_linetype_manual(values=1:3, name="Tradeoff (n)") 
```

@@

```{r Same resource habitat, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.cap="Effect of character divergence on total attack rates (A) and resource abundances (B) when both resources are found in the same habitat."}
Mac_attack <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=aii+aij, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=a11+a12, fill=Competitor), shape=21, size=5, inherit.aes = FALSE) +
  ylab(expression(a[ii]+a[ij])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_linetype_manual(values=1:3, name="Tradeoff (n)") +
  scale_fill_manual(values=c("white","black"))

@@

Mac_plots <- plot_grid(Mac_attack + theme(legend.position = "none"), 
                       Mac_resources + theme(legend.position = "none"), 
                       nrow=1, labels = "AUTO")

@@

#Mac_plot_legend <- 
plot_grid(Mac_plots, Mac_legend, ncol=2, rel_widths = c(3,0.3), align='hv')

#Mac_title <- ggdraw() + draw_label("Resources in same habitat", fontface='bold')
#plot_grid(Mac_title, Mac_plot_legend, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins
```

@@

```{r Distinct resource habitats, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.cap="Effect of character divergence on total attack rates (A) and resource abundances (B) when each resource is found in a different habitat."}
LS_attack <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=wii*aii+wij*aij, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=w11*a11+w12*a12, fill=Competitor), shape=21, size=5, inherit.aes = FALSE) +
  ylab(expression(w[ii]*a[ii]+(1-w[ii])*a[ij])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_linetype_manual(values=1:3, name="Tradeoff (n)") +
  scale_fill_manual(values=c("white","black"))

@@

LS_plots <- plot_grid(LS_attack + theme(legend.position = "none"), 
                      LS_resources + theme(legend.position = "none"), 
                      nrow=1, labels = "AUTO")

@@

#LS_plot_legend <- 
plot_grid(LS_plots, LS_legend, nrow=1, rel_widths = c(3,0.3))

#LS_title <- ggdraw() + draw_label("Resources in different habitats", fontface='bold')
#plot_grid(LS_title, LS_plot_legend, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins
```

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Effect of character divergence on total attack rates (A) and resource abundances (B) when consumers have a more realistic functional response. Note also that each resource is found in a different habitat."}
McCann_attack <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=wii*aii+wij*aij, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(McCann_displacements, Time=="End"), aes(x = a11/(a11+a12), y=w11*a11+w12*a12, fill=Competitor), shape=21, size=5, inherit.aes = FALSE) +
  ylab(expression(w[ii]*a[ii]+(1-w[ii])*a[ij])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_linetype_manual(values=1:3, name="Tradeoff (n)") +
  scale_fill_manual(values=c("white","black"))

@@

McCann_plots <- plot_grid(McCann_attack + theme(legend.position = "none"), 
                          McCann_resources + theme(legend.position = "none"), 
                          nrow=1, labels = "AUTO")

@@

#McCann_plot_legend <- 
plot_grid(McCann_plots, McCann_legend, nrow=1, rel_widths = c(3,0.3))

#McCann_title <- ggdraw() + draw_label("Resources in different habitats & realistic functional response", fontface='bold')
#plot_grid(McCann_title, McCann_plot_legend, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins
```

Of the prior models of character displacement, only Lawlor and Smith (1976) examine the effect of character divergence on the ecological stability of the system. They conducted a detailed theoretical analysis of three different metrics of stability: (i) stable coexistence of consumers, defined as the mutual invasibility of each consumer; (ii) global stability of consumers, defined as the continued coexistence of both consumers when resources dynamics are perturbed; and (iii) resilience, defined as the rate at which the four-species community returns to equilibrium after a small perturbation to species abundances. Note that the first two metrics correspond to the stable coexistence of consumers, while the third metric measures the stability of the entire system. Lawlor and Smith found that character divergence enhanced the stable coexistence of consumers, yet had a small negative effect of on food-web resilience (which they downplay in their interpretation).

Lawlor and Smith's work gave fundamental insight to the effect of character divergence on ecological stability; however, none of their stability analyses compared systems with and without a competitor. Such comparisons are necessary for inferring the effects of ecological character displacement, a point that has been made clear in the criteria to demonstrate character displacement (Schluter and McPhail 1992; Schluter 2000). In fact, the first two metrics of stability can only be assessed when both consumers are present, because it interprets the effect of divergence in the presence --- not absence --- of a competitor. Food-web reslience, however, can be analyzed in the presence and absence of a competitor, and thus provides a useful benchmark for exploring the effects of character displacement on food-web stability.

To gain analytical insight, we examined the conditions governing the local stability of the four-species food web when consumers have a more realistic functional response (McCann 2005). In the Appendix, we show that the system will transition from being locally stable to a limit cycle under the following conditions:

$$w_{ii}a_{ii}+(1-w_{ii})a_{ij} < \frac{e+hm}{hK(e-hm)}$$

Previously, we showed that character displacement always increases the effective attack rate of consumers, regardless of the the shape of the tradeoff. Thus, character divergence will push system toward the boundary of stability where the system transitions from a stable equilibrium to a limit cycle. This result supports the general tendency observed by Lawlor and Smith's numerical examination of local stability when consumers exhibited linear functional responses.

Still, this does not compare the resilience of food-webs in the presence and absence of a competitor. To explore this, we plot the results from the same simulations for resource abundances, but now stability is on the y-axis

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Effect of character divergence on food-web resilience across different foraging scenarios and tradeoffs in attack rates."}
Stability_Mac <- cbind.data.frame(tradeoff_df, MacArthur_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=Competitor), shape=21, size=5, inherit.aes = FALSE) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression(-1%*%"Re"(lambda[max]))) +
  scale_linetype_manual(values=1:3, name="Tradeoff (n)") +
  scale_fill_manual(values=c("white","black"))

Stability_LS <- cbind.data.frame(tradeoff_df, LawlorSmith_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=Competitor), shape=21, size=5, inherit.aes = FALSE) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression(-1%*%"Re"(lambda[max]))) +
  scale_linetype_manual(values=1:3, name="Tradeoff (n)") +
  scale_fill_manual(values=c("white","black"))

Stability_McCann <- cbind.data.frame(tradeoff_df, McCann_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, linetype = factor(n))) + 
  geom_line() +
  geom_point(data = filter(McCann_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=Competitor), shape=21, size=5, inherit.aes = FALSE) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression(-1%*%"Re"(lambda[max]))) +
  scale_linetype_manual(values=1:3, name="Tradeoff (n)") +
  scale_fill_manual(values=c("white","black"))

Stability_plots <- plot_grid(Stability_Mac + theme(legend.position = "none"), 
                             Stability_LS + theme(legend.position = "none"),
                             Stability_McCann + theme(legend.position = "none"),
                             nrow=1, labels = "AUTO")

Stability_legend <- get_legend(Stability_Mac)

#Stability_plot_legend <- 
plot_grid(Stability_plots, Stability_legend, nrow=1, rel_widths = c(3,0.3))
```

@@
